---
title: "1_GWAS_Ancestry_Comparison"
author: "Jonathan Chan"
date: "13/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

# Comparison of Number of GWAS Participants by Self-Reported Ancestry

## Aim

The aim of this code is to output a comparative plot of the number of GWAS particpant by their ancestry to outline the ancestral disparities in population genetics studies. A reference plot can be found in Martin et al. (2019).

## Method

The GWAS Catalog (NHGRI-EBI) will be used as source of data for current GWAS studies - both published and unpublished.
Hence, their V1.0.3 ancestry dataset will be used as the base dataset: https://www.ebi.ac.uk/gwas/api/search/downloads/ancestries_new

The output .tsv file will be imported as a tibble (Tidyverse) into R and summary statistics across all studies will be carried out to derive a summary statistics table. This will then be transformed into a summative plot.

```{r Data_Import}
# Below code downloads the latest source data from GWAS catalog and imports it as a tibble into R and discarding columns which aren't required.
Sys.time()
download.file('https://www.ebi.ac.uk/gwas/api/search/downloads/ancestries_new', 'ancestry_GWAS.tsv')
base_data <- read_tsv('ancestry_GWAS.tsv') %>%
  filter(!str_detect(`BROAD ANCESTRAL CATEGORY`,'.?NR.?'))
  select(1,3,4,5,6,8,9)
```

The broad ancestral categories that will be compared  will be:
* European
* East Asian
* South/Other Asian
* African
* Hispanic/Latino
* Greater Middle Eastern
* Oceanian
* Other
* Multiple

```{r Ancestry_Cleaner}
# Below code filters the dataset into individual tibbles corresponding to studies for each of the broad ancestral categories, excluding the non-reported multiple ancestry categories and simplifying labels to the broad ancestral categories, and then recombines them into a simplified tibble.

european <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('European'))
e_asian <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('East Asian'))
s_asian <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('South Asian','South East Asian','Central Asian', 'Asian unspecified'))
s_asian$`BROAD ANCESTRAL CATEGORY` <- 'South/Other Asian'
  
african <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('African American or Afro-Caribbean','African unspecified', 'Sub-Saharan African', 'African American or Afro-Caribbean, African unspecified', 'African unspecified, African American or Afro-Caribbean', 'Sub-Saharan African, African unspecified', '	
African unspecified, Sub-Saharan African', 'Sub-Saharan African, African American or Afro-Caribbean'))
african$`BROAD ANCESTRAL CATEGORY` <- 'African'

hispanic <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('Hispanic or Latin American'))
gme <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('Greater Middle Eastern (Middle Eastern, North African or Persian)'))
gme$`BROAD ANCESTRAL CATEGORY` <- 'Greater Middle Eastern'

oceanian <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('Oceanian','Aboriginal Australian'))
oceanian$`BROAD ANCESTRAL CATEGORY` <- 'Oceanian'

other <- filter(base_data, `BROAD ANCESTRAL CATEGORY` %in% c('Native American','Other','Other admixed ancestry'))
other$`BROAD ANCESTRAL CATEGORY` <- 'Other'

multiple <- filter(base_data, str_detect(`BROAD ANCESTRAL CATEGORY`, ',')) %>%
  filter(!(`BROAD ANCESTRAL CATEGORY`  %in% c('Greater Middle Eastern (Middle Eastern, North African or Persian)','African American or Afro-Caribbean, African unspecified', 'African unspecified, African American or Afro-Caribbean', 'Sub-Saharan African, African unspecified', '	
African unspecified, Sub-Saharan African','Sub-Saharan African, African American or Afro-Caribbean')))
multiple$`BROAD ANCESTRAL CATEGORY` <- 'Multiple'

total_simple_base_data <- bind_rows(european, e_asian, s_asian, african, hispanic, gme, oceanian, other, multiple) %>%
  arrange(DATE)
```

```{r Comparative_Plot}
# The following code plots the comparative time series plot with y-axis

```

